---
title: "[CSDE 502 Winter 2021](http://staff.washington.edu/phurvitz/csde502_winter_2021) Assignment 2"
author: "answer key"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output: 
    html_document:
        number_sections: false
        self_contained: true
        code_folding: show
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
    pdf_document:
        number_sections: true
        toc: true
        fig_cap: yes
        keep_tex: yes
urlcolor: blue        
---

<!--these following settings control width and also make a DIV tag for black outlines for figures-->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)

# captions
library(captioner)
table_nums <- captioner(prefix = "Table")
figure_nums <- captioner(prefix = "Figure")

# for having DIV tags
knitr::knit_hooks$set(class = function(before, options, envir) {
  if(before){
    sprintf("<div class = '%s'>", options$class)
  }else{
    "</div>"
  }
})

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

<!-- for a wide presentation -->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<!-- a DIV for black outline, good for figures-->
<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

# Q. 5.1
__Present three tables that have all of the original variables, one for each species of iris.__

The `iris` data frame can presented as three tables based on species using several different methods.

## Brute force
The brute force method performs each step manually.

Unique species: 

```{r}
# unique values of species
(spp <- unique(iris$Species))

# make each data frame
setosa <- iris %>% filter(Species == "setosa")
versicolor <- iris %>% filter(Species == "versicolor")
virginica <- iris %>% filter(Species == "virginica")

# print each data frame
kable(setosa, caption = "setosa") %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left") 

kable(versicolor, caption = "versicolor") %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left") 

kable(virginica, caption = "virginica") %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F,
                  position = "left") 
```

## Data-driven grouping `group_*()` functions in `dplyr`
Some automation can be applied. While this method took more lines of code to write, it did not require any interactive identification or naming of species, which could easily result in typographical errors for a longer list of values.

Because the code chunk generates printable HTML, use the the code chunk option `results='as-is'`.

```{r, results='asis'}
# group by species
spp_g <- iris %>% 
    group_by(Species)

# split
spp_gs <- spp_g %>% 
    group_split()

# unique species
spp <- spp_g %>% 
    group_keys() %>% 
    pull(Species)

# loop
for(i in 1:length(spp)){
    # species
    spp_i <- spp[i]
    # print table
    print(kable(spp_gs[i][[1]], caption = spp[i]) %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left"))
}
```

# Q. 5.2
__Present a table with only sepal length and sepal width for spp. virginica that also has columns indicating whether the sepal length and sepal width are greater than the mean for this species.__

The mean of the different species' sepal length and witdth, for validation:

```{r}
iris %>% 
    group_by(Species) %>% 
    summarise(mean_sepal_length = mean(Sepal.Length),
              mean_sepal_width = mean(Sepal.Width)) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left")
```

The following table restricts to species _virginica_ and gets only sepal length and width, and indicates whether the sepal length and width are greater than the mean for that species. It is not ncessary, and it is not advised to use a static number in the logical test for `<` or `>` compared to the mean.

```{r}
i <- iris %>% 
    filter(Species == "virginica") %>% 
    select(Species, Sepal.Length, Sepal.Width) %>% 
    mutate("sepal length > mean" = Sepal.Length > mean(Sepal.Length),
           "sepal width > mean" = Sepal.Width > mean(Sepal.Width))

kable(i) %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left")
```

# Q. 5.3
__Create a new variable that classifies health, stratified at the break between "(2) Very good" and better versus "(3) Good" and worse. Make sure you explicitly handle missing or unknown values.__

Looking at the structure of the data ([week 2]()http://staff.washington.edu/phurvitz/csde502_winter_2021/week2.html#tidyverse), we see that the "general health" variable is named `h1gh1` and has 7 levels:

```{r, warning=FALSE}
dat <- readstata13::read.dta13("http://staff.washington.edu/phurvitz/csde502_winter_2021/data/AHwave1_v1.dta")

write.csv(levels(dat$h1gh1), row.names = FALSE)
```

There are a few different ways one could stratify the variable; this example will use the `case_when()` function one case at a time. Any values that were not between Poor and Excellent were treated as NA:

```{r}
# stratify health
dat %<>%  
    mutate(health = 
    case_when(
        h1gh1 == "(1) Excellent" ~ "high",
        h1gh1 == "(2) Very good" ~ "high",
        h1gh1 == "(3) Good" ~ "low",
        h1gh1 == "(4) Fair" ~ "low",
        h1gh1 == "(5) Poor" ~ "low",
        TRUE ~ as.character(NA)
    ))
```

We might simplify this a bit by creating an ordered factor variable, to use numerical comparisons. 

```{r}
# ordered factor; use fct_rev to establish the correct ordering where better health ranks higher
dat %<>% 
    mutate(h1gh1 = fct_rev(as.ordered(h1gh1)))

# confirm ordering
unique(dat$h1gh1)
```

Stratificaton can then be done using factor levels as ordinals. Additionally, the new variable can be specified as an ordered factor.

```{r}
# stratify health; first we need to catch the "don't know" and "refused" as NAs
dat %<>%  
    mutate(health = 
        case_when(
            h1gh1 <= "(6) Refused" ~ as.character(NA),
            h1gh1 >  "(3) Good" ~ "Very good or better",
            h1gh1 <= "(3) Good" ~ "Good or worse"
        ) %>% 
            factor(levels = c("Very good or better", "Good or worse"))
    )
```

# Q. 5.4
__Tabulate (count and percent) this new variable for all respondents.__

```{r}
tabhealth <- dat %>% 
    group_by(health) %>% 
    summarise(n = n()) %>% 
    mutate("%" = round(n / sum(n) * 100, 2))

kable(tabhealth) %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left")
```

# Q. 5.5
__Tabulate (count and percent) this new variable for those who self-identified as White versus those who self-identified as African American.__

## Treating each race separately
There are two variables, from distinct questions dealing with race: h1gi6a (S1Q6A RACE-WHITE-W1), and h1gi6b (S1Q6B RACE-AFRICAN AMERICAN-W1), that are not mutually exclusive, so each variable is treated separately. Some people may have self-reported being both White and African-American.

```{r}
# tabulate by White
tabhealth_white <- dat %>% 
    filter(h1gi6a == "(1) Marked") %>% 
    group_by(health) %>% 
    summarise(n = n(), .groups = "drop_last") %>% 
    mutate("%" = round(n / sum(n) * 100, 2))

# tabulate by African American
tabhealth_afram <- dat %>% 
    filter(h1gi6b == "(1) Marked") %>% 
    group_by(health) %>% 
    summarise(n = n(), .groups = "drop_last") %>% 
    mutate("%" = round(n / sum(n) * 100, 2))

# column-bind and remove the second "health" column
x <- cbind(tabhealth_white, tabhealth_afram) 
x[,4] <- NULL

# print the table with better printed names
kable(x, col.names = c("health", "n", "%", "n", "%")) %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, 
                  position = "left") %>% 
    add_header_above(c(" " = 1, "White" = 2, "African American" = 2))
```
## Treating each race simultaneously
Because these variables are not mutually exclusive, in order to create a single mutually excluse variable (including "other"), rules need to be established. For this exercise we will make these assumptions:

1. "White" will refer to those who reported being White and _not_ being African-American
1. "African American" will refer to those who reported being African-American and _not_ being White
1. "other" will be any cases that do not match the previous conditions.

```{r}
# mutate using case_when() for race; also make race an ordered factor, then group_by() and summarize()
tabhealth_w_aa_o <- dat %>% 
    mutate(race = 
               case_when(
                  h1gi6a == "(1) Marked" & h1gi6b == "(0) Not marked" ~ "White",
                  h1gi6b == "(1) Marked" & h1gi6a == "(0) Not marked" ~ "African-American",
                  TRUE ~ "other") %>% 
               factor(levels = c("White", "African-American", "other"))) %>% 
    group_by(race, health) %>% 
    summarise(n = n(), .groups = "drop_last") %>% 
    mutate("%" = round(n / sum(n) * 100, 2))

# print the table with better printed names
# ungroup() so we can drop the first column
tabhealth_w_aa_o %>% 
    ungroup() %>% 
    select(-1) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = 
                      c("striped", "hover", "condensed", "responsive"),
                  full_width = F, 
                  position = "left") %>% 
    pack_rows("White", 1, 3) %>% 
    pack_rows("African American", 4, 6) %>% 
    pack_rows("other", 7, 9)
```


<h1>Source code for this document</h1>
```{r comment=''}
cat(readLines("csde502_2021_assignment02_answers.Rmd"), sep = '\n')
```
