---
title: "Assignment 6 example answer key"
author: "Phil Hurvitz | [e-mail](mailto:phurvitz@uw.edu) | [www](http://gis.washington.edu/phurvitz)"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output: 
    bookdown::html_document2:
        number_sections: true
        self_contained: true
        code_folding: show
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
urlcolor: blue
---

<!--these following settings control width and also make a DIV tag for black outlines for figures-->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

```{r setup, message=FALSE, warning=FALSE}
# key setups
library(knitr) # knitr for kable tables
library(kableExtra) # pretty tables
library(ggplot2)
library(tidyverse)
library(pdftools)
library(labelled)

# captions
library(captioner)
table_nums <- captioner(prefix = "Table")
figure_nums <- captioner(prefix = "Figure")

# for having DIV tags
knitr::knit_hooks$set(class = function(before, options, envir) {
    if(before){
        sprintf("<div class = '%s'>", options$class)
    }else{
        "</div>"
    }
})

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

<!-- for a wide presentation -->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<!-- a DIV for black outline, good for figures-->
<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

<h2>Q. 2.1</h2>
__In what pdf document is the documentation for the race items for the Wave I In-Home questionnaire?__

The following function matches strings in PDF files. Because we are searching for the _word_ race, we use the regular expression `\\brace\\`. The second backslash is needed in R to specify this is a literal backslash character.

The R help on regular expressions shows:

>The caret ^ and the dollar sign $ are metacharacters that respectively match the empty string at the beginning and end of a line. The symbols \< and \> match the empty string at the beginning and end of a word. The symbol \b matches the empty string at either edge of a word, and \B matches the empty string provided it is not at an edge of a word. (The interpretation of ‘word’ depends on the locale and implementation: these are all extensions.)

The results show the regular expression that was used, the page number on which the regular expression was found, and that the pattern match was case insensitive (i.e., search for `race`, `Race`, `RACE`, etc.).

```{r}
# a function to get matching strings in a PDF, ignore case
f_pdf_str_match <- function(x, pat, ignore.case = TRUE){
    # convert the PDF to text
    mytext <- pdf_text(x)
    # pattern
    if(ignore.case){
        mypat <- regex(pat, ignore_case = TRUE)
    } else {
        mypat <- pat
    }
    # match strings = pages
    pages <- str_which(string = mytext, pattern = mypat)
    if(length(pages) == 0){
        return(data.frame(fname = basename(x), pat, page_num = as.integer(NA), ignore.case))
    }
    # create a data frame
    data.frame(fname = basename(x), pat, page_num = pages, ignore.case)
}

# a list of my PDFs
mypdfs <- list.files(path = "../data/metadata/Wave1_InHome_Codebooks", pattern = "*.pdf$", full.names = TRUE)

# an empty data frame
x <- NULL

# run each one
for(i in mypdfs){
    x <- rbind(x, f_pdf_str_match(i, "\\brace\\b", ignore.case = TRUE))
}

# ignore NAs
x %>% filter(!is.na(page_num))
```


<h2>Q. 3.2</h2>	
__Using questions 6 and 8 in `inh01pub.pdf`, create a new variable named "race" that uses recoded values (white = 1; black/African American = 2; American Indian = 3; Asian/Pacific Islander = 4; other = 5; unknown/missing = 9).__

Before answering the question, we should understand the structure of responses to questions 6 and 8.

The variables and variable labels for question 6 are shown in Table \@ref(tab:q6).

```{r q6}
# read the data set
dat <- haven::read_dta("http://staff.washington.edu/phurvitz/csde502_winter_2021/data/AHwave1_v1.dta")
dat %>% 
    select(str_subset(names(dat), "^h1gi6|^h1gi8")) %>% 
    map(~attributes(.)$label) %>% 
    unlist() %>% 
    data.frame(label = .) %>% 
    kable(caption = "Question 6 variables") %>% 
    kable_styling(full_width = FALSE, position = "left")
```

Response values for question 6 are shown in Table \@ref(tab:race6). A value of `1` indicates that the respondent confirmed having the particular racial designation in question.

```{r race6}
attributes(dat$h1gi6a)$labels %>% data.frame(value = .) %>% 
    kable(caption = "Values for question 6") %>% 
    kable_styling(full_width = FALSE, position = "left")
```

For question 8, it is a single race question with responses shown in Table \@ref(tab:race8). A response of `7` indicates that they only chose one race in question 6. So for example, if they marked `h1gi6c == 1` and `h1gi8 == 7`, they classified themselves as only "American Indian/Native American". 


```{r race8}
# american indian/native american for text below
# responded 1 to question 6c but included other races as well
n_ai_q6c <- dat %>% filter(h1gi6c == 1) %>% nrow()
# responded 3 to question 8
n_ai_q8 <- dat %>% filter(h1gi8 == 3) %>% nrow()
# responded 1 to question 6c and responded 7 to question 8 (marked only one race in q6)
n_ai_q6c_q8 <- dat %>% filter(h1gi6c == 1 & h1gi8 == 7) %>% nrow()

attributes(dat$h1gi8)$labels %>% data.frame(value = .) %>% 
    kable(caption = "Responses for question 8") %>% 
    kable_styling(full_width = FALSE, position = "left")
```

Let's unpack this a bit. For those who responded `h1gi6c` = `1`, that is, consider themselves American Indian/Native American (AI/NA, ``r attributes(dat$h1gi6c)$label``), there were `r n_ai_q6c` positive responses. However, that included persons who also marked being more than one race for other parts of questions 6. To get those who considered themselves _only_ AI/NA, we need to find those with `h1gi8 == 3` (i.e., ``r attributes(dat$h1gi8)$labels[3] %>% names()``), a count of `r n_ai_q8`, plus those who had a combination of `h1gi6c == 1` and `h1gi8 == 7` (i.e., ``r attributes(dat$h1gi8)$labels[7] %>% names()``), a count of `r n_ai_q6c_q8`, for a total of `r n_ai_q8 + n_ai_q6c_q8`.

In the following code, the new variable will be created using `mutate()` (to create the new column) `case_when()` (to handle different cases), `Hmisc::label()` (to label the column), and `factor()` (to label the numerical values). Additionally, we create the `labels` attribute of the `race` column to be compliant with the attribute value naming scheme of the other columns.

```{r case_when}
# a function to label a column
labelled <- function(x, label) {
    Hmisc::label(x) <- label
    x
}

dat %<>% 
    mutate(race = case_when(
        # the cases cover q. 6 and 8
        # for example, the first test is 
            # h1gi8 = 1 OR h1gi8 = 7 and h1gi6a = 1 (i.e., White)
        h1gi8 == 1 | (h1gi8 == 7 & h1gi6a == 1) ~ 1,
        h1gi8 == 2 | (h1gi8 == 7 & h1gi6b == 1) ~ 2,
        h1gi8 == 3 | (h1gi8 == 7 & h1gi6c == 1) ~ 3,
        h1gi8 == 4 | (h1gi8 == 7 & h1gi6d == 1) ~ 4,
        h1gi8 == 5 | (h1gi8 == 7 & h1gi6e == 1) ~ 5,
        # missing
        h1gi8 %in% c(6, 8, 9) ~ 9)
    ) %>% 
    # anything that was NA becomes 9
    mutate(
        race = replace_na(race, 9) %>% labelled("race as mutually exclusive")
    ) %>% 
    # create factor labels
    mutate(race = factor(race, labels = c(
        "white", 
        "black/African American", 
        "American Indian", 
        "Asian/Pacific Islander", 
        "other", 
        "unknown/missing"
)))

# column labels attribute
attributes(dat$race)$labels <- c(
        "white" = 1, 
        "black/African American" = 2, 
        "American Indian" = 3, 
        "Asian/Pacific Islander" = 4, 
        "other" = 5, 
        "unknown/missing" = 9)
```

The tabulation of subject counts by the new variable is shown in Table \@ref(tab:racetab).

```{r racetab}
dat %>% 
    group_by(race) %>% 
    summarise(n = n()) %>% 
    kable(caption = "Tabulation of new single race variable") %>% 
    kable_styling(full_width = FALSE, position = "left")
```

<h2>Q. 3.3</h2>
__Rename the variables, and update variable labels using Table 1 as a guide and save the data frame as the file as `AHwave1_v2.rds`. Use a single R code file for your edits to the data file.__

There are at least two approaches to this.

The first approach here will use a single dplyr statement, through the use of `mutate()` (to create new variables with specified names), `Hmisc::label()` (to label those new variables), and `select(-c(...))` (to drop columns that are redundant from renaming/labeling).

```{r}
# a function to label a column
labelled <- function(x, label) {
    Hmisc::label(x) <- label
    x
}

# rename variables
dat2 <- dat %>% 
  mutate(
        aid = aid %>% labelled("unique case (student) identifier"),
        imonth = imonth %>% labelled("month interview completed"),
        iday = iday  %>% labelled("day interview completed"),
        iyear = iyear %>% labelled("year interview completed"),
        bio_sex = bio_sex  %>% labelled("interviewer confirmed sex"),
        bmonth = h1gi1m %>% labelled("birth month"),
        byear = h1gi1y %>% labelled("birth year"),
        hispanic = h1gi4 %>% labelled("Hispanic/Latino"),
        white = h1gi6a %>% labelled("race white"),
        black = h1gi6b %>% labelled("race black or African American"),
        AI = h1gi6c %>% labelled("race American Indian or Native American"),
        asian = h1gi6d %>% labelled("race Asian or Pacific Islander"),
        raceother = h1gi6e %>% labelled("race other"),
        onerace = h1gi8 %>% labelled("one category best describes racial background"),
        observedrace = h1gi9 %>% labelled("interviewer observed race"),
        health = h1gh1 %>% labelled("how is your health")) %>% 
    select(-c(h1gi1m, h1gi1y, h1gi4, h1gi6a, h1gi6b, h1gi6c, h1gi6d, h1gi6e, h1gi8, h1gi9, h1gh1))
```

The second approach uses `rename()` to rename a set of columns _en masse_:

```{r}
dat3 <- dat %>% 
  rename(bmonth = h1gi1m,
         byear = h1gi1y,
         hispanic = h1gi4,
         white = h1gi6a,
         black = h1gi6b,
         AI = h1gi6c,
         asian = h1gi6d,
         raceother = h1gi6e,
         onerace = h1gi8,
         observedrace = h1gi9,
         health = h1gh1)
```

The third (base R) approach uses `grep()` to find the position of the column name and reassign.

```{r}
# copy the data frame
dat4 <- dat
# find the named columns and rename them
colnames(dat4)[grep("^h1gi1m$", colnames(dat4))] <- "bmonth"
colnames(dat4)[grep("^h1gi1y$", colnames(dat4))] <- "byear"
colnames(dat4)[grep("^h1gi4$", colnames(dat4))] <- "hispanic"
colnames(dat4)[grep("^h1gi6a$", colnames(dat4))] <- "white"
colnames(dat4)[grep("^h1gi6b$", colnames(dat4))] <- "black"
colnames(dat4)[grep("^h1gi6c$", colnames(dat4))] <- "AI"
colnames(dat4)[grep("^h1gi6d$", colnames(dat4))] <- "asian"
colnames(dat4)[grep("^h1gi6e$", colnames(dat4))] <- "raceother"
colnames(dat4)[grep("^h1gi8$", colnames(dat4))] <- "onerace"
colnames(dat4)[grep("^h1gi9$", colnames(dat4))] <- "observedrace"
colnames(dat4)[grep("^h1gh1$", colnames(dat4))] <- "health"
```

For both the second and third method, column labels can be handled using `labelled::var_label()`:

```{r labels}
# change labels as specified
var_label(dat3$aid) <- "unique case (student) identifier"
var_label(dat3$imonth) <- "month interview completed"
var_label(dat3$iday) <- "day interview completed"
var_label(dat3$iyear) <- "year interview completed"
var_label(dat3$bio_sex) <- "interviewer confirmed sex"
var_label(dat3$bmonth) <- "birth month"
var_label(dat3$byear) <- "birth year"
var_label(dat3$hispanic) <- "Hispanic/Latino"
var_label(dat3$white) <- "race white"
var_label(dat3$black) <- "race black or African American"
var_label(dat3$AI) <- "race American Indian or Native American"
var_label(dat3$asian) <- "race Asian or Pacific Islander"
var_label(dat3$raceother) <- "race other"
var_label(dat3$onerace) <- "one category best describes racial background"
var_label(dat3$observedrace) <- "interviewer observed race"
var_label(dat3$health) <- "how is your health"

var_label(dat4$aid) <- "unique case (student) identifier"
var_label(dat4$imonth) <- "month interview completed"
var_label(dat4$iday) <- "day interview completed"
var_label(dat4$iyear) <- "year interview completed"
var_label(dat4$bio_sex) <- "interviewer confirmed sex"
var_label(dat4$bmonth) <- "birth month"
var_label(dat4$byear) <- "birth year"
var_label(dat4$hispanic) <- "Hispanic/Latino"
var_label(dat4$white) <- "race white"
var_label(dat4$black) <- "race black or African American"
var_label(dat4$AI) <- "race American Indian or Native American"
var_label(dat4$asian) <- "race Asian or Pacific Islander"
var_label(dat4$raceother) <- "race other"
var_label(dat4$onerace) <- "one category best describes racial background"
var_label(dat4$observedrace) <- "interviewer observed race"
var_label(dat4$health) <- "how is your health"
```

Note that for the first method, the order of the columns is changed. This may cause problems for any code that uses hard-coded column positions rather than column names.

```{r}
# save 
write_rds(x = dat2, file = "C:/users/phurvitz/csde502/data/AHwave1_v1_tidy1.rds")
write_rds(x = dat3, file = "C:/users/phurvitz/csde502/data/AHwave1_v1__tidy2.rds")
write_rds(x = dat4, file = "C:/users/phurvitz/csde502/data/AHwave1_v1.rds")
```
