---
title: "Assignment 7 example answer key"
author: "Phil Hurvitz | [e-mail](mailto:phurvitz@uw.edu) | [www](http://gis.washington.edu/phurvitz)"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output: 
    bookdown::html_document2:
        number_sections: true
        self_contained: true
        code_folding: hide
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
urlcolor: blue
---

<!--these following settings control width and also make a DIV tag for black outlines for figures-->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

```{r setup, message=FALSE, warning=FALSE}
# key setups
library(knitr) # knitr for kable tables
library(kableExtra) # pretty tables
library(ggplot2)
library(tidyverse)
library(pdftools)
library(labelled)
library(magrittr)

# captions
library(captioner)
table_nums <- captioner(prefix = "Table")
figure_nums <- captioner(prefix = "Figure")

# for having DIV tags
knitr::knit_hooks$set(class = function(before, options, envir) {
    if(before){
        sprintf("<div class = '%s'>", options$class)
    }else{
        "</div>"
    }
})

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

<!-- for a wide presentation -->
<style type="text/css">
.main-container {
  max-width: 2400px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

<!-- a DIV for black outline, good for figures-->
<style>
.outline {
   background-color: #000000;
   padding:1px;
   display: inline-block;
}
</style>

This document presents some possible answers for Assignment 7.

<h2>1.</h2>
__Generate a new data frame from the full public Add Health data set (http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip) that consists of a subset of at least four of the original columns. Do not use any of the variables used as examples in Lesson 7.__

In this code chunk, the file is downloaded, unzipped, read in as a data frame, and limited to five columns with a `select()`.

```{r}
myUrl <- "http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip"
# zipfile in $temp
zipfile <- file.path(Sys.getenv("TEMP"), basename(myUrl))
# dta file in $temp
dtafile <- tools::file_path_sans_ext(zipfile)
# check if the dta file exists
if(!file.exists(dtafile)){
    # if the dta file doesn't exist, check for the zip file
    # check if the zip file exists, download if necessary
    if(!file.exists(zipfile)){
        curl::curl_download(url = myUrl, destfile = zipfile)
    }
    # unzip the downloaded zip file
    unzip(zipfile = zipfile, exdir = Sys.getenv("TEMP"))
}

# if the data set has not been read, read it in 
if(!exists("ahcomplete")){
    ahcomplete <- haven::read_dta(dtafile)
}
# lowercase column names
colnames(ahcomplete) %<>% str_to_lower()

# create a subset of columns
dat <- ahcomplete %>% 
    select(aid, 
           s2,
           s10a,
           s46c,
           s47,
           s62m)
# new column names
newcolnames <- c("aid", "sex", "mathgrade", "homework", "tvtime", "likemyself")
```


<h3>1.a</h3>
__At least one of the variables should be able to stratify the respondents into meaningful analytic groups.__

All of the variables other than `aid` could be used to stratify. The variables I selected are ``r dat %>% names() %>% str_c(collapse = ", ")``, renamed to ``r newcolnames %>% str_c(collapse = ", ")``

```{r}
# set the column names
names(dat) <- newcolnames
```

<h3>1.b</h3>
__The data frame should have a “label” attribute that provides a brief but informative description of the table.__

```{r}
attributes(dat)$label <- "A subset of Add Health public data.
Selected variables were:
s2, s10a, s46c, s47, and s62m. 
The variables were renamed to:
aid, sex, mathgrade, homework, tvtime, and likemyself."
```

The descriptive label is:

```{r, comment=""}
# show the labels
cat(attributes(dat)$label)
```

<h3>1.c</h3> 
__The columns should be formatted as factor variables with proper value labels and ordering if applicable.__

The following code converts the variables to factors.

```{r}
# save a previous version for comparison
datorig <- dat

# a function to convert a column with value attribute labels to a factor
f_fct <- function(x){
    factor(x, labels = attributes(x)$labels %>% names(), ordered = TRUE)
}

# mutate all columns other than the first (aid), using the f_fct function
#   to convert the columns to factors
dat %<>%
    mutate_at(.vars = -1,
              .funs = f_fct
    )
```

Here we see the first few records of the data frame before and after converting variables to factors (\@ref(tab:datorig), \@ref(tab:datfact), respectively).

```{r datorig}
# show the first few records
datorig %>% 
    head() %>% 
    kable(caption = "Raw data") %>% 
    kable_styling(full_width = FALSE, position = "left")
```

```{r datfact}
# show the first few records
dat %>% 
    head() %>% 
    kable(align = rep("r", ncol(dat)), caption = "Data converted to factors") %>% 
    kable_styling(full_width = FALSE, position = "left")
```


<h3>1.d</h3> 
__The columns should have informative attributes as you see fit.__

I have used the labels from the original data set to label the new variables, which are listed below.

```{r}
# column label
labs <- datorig %>% 
    map(~ attributes(.)$label) %>% 
    unlist()

# run a loop to label the columns
for(i in 1:ncol(dat)){
    # get the label
    mylab <- labs[i]
    # get the column name
    cname <- names(dat)[i]
    # formulate a command
    mycmd <- sprintf("attributes(dat$%s)$label <- '%s'", cname, mylab)
    # run the command
    eval(parse(text = mycmd))
}

# print the labels
dat %>% 
    map(~ attributes(.)$label)
```


<h2>2</h2>
__The code should save the data frame as an RDS file in the $TEMP location, which can be specified in R as Sys.getenv("TEMP").__

The following code saves the data set as an RDS file (``file.path(tmpdir, "phurvitz_assn8.Rmd")``).

```{r}
# save the data set
tmpdir <- Sys.getenv("TEMP")
saveRDS(object = dat, file = file.path(tmpdir, "phurvitz_assn8.Rmd"))
```


<h2>3</h2>
__Create some frequency tables:__

<h3>3.a</h3>
__Create a frequency table from each variable, using both counts and percentages.__

The following code runs a loop over each column other than `aid` to create and print a tabulation.

```{r, results='asis'}
# create a tabulation for each column other than aid
for(i in 2:ncol(dat)){
    # column name
    cname <- names(dat)[i]
    
    # tabulate
    ftab <- dat %>% 
        group_by(dat[,i]) %>% 
        summarise(n = n()) %>% 
        mutate(`%` = (n / sum(n) * 100) %>% round(1)) %>% 
        kable() %>% 
        kable_styling(full_width = FALSE,
                      position = "left") %>% 
        print()
}
```


<h3>3.b</h3>
__Create a frequency table based on at least two variables, also with counts and percentages.__

The following tables present tabulations of grades in math by time spent watching TV. Table \@ref(tab:mt) presents percentage sums by grade, and \@ref(tab:tm) presents percentage sums by TV time.

```{r mt}
dat %>% 
    filter(mathgrade <= "(4) D or lower" &
               tvtime < "(9) Multiple response") %>% 
    group_by(mathgrade, tvtime) %>% 
    summarise(n = n()) %>% 
    mutate(pct = (n / sum(n) * 100) %>% round(1),
           `n (%)` = str_c(n, " (", pct, ")", sep = "")) %>% 
    select(mathgrade, tvtime, `n (%)`) %>% 
    spread(mathgrade, `n (%)`) %>% 
    kable(caption = "Tabulation of grades in math and TV time, percent total by grade") %>% 
    kable_styling(full_width = FALSE,
                  position = "left") %>% 
    add_header_above(header = c(" " = 1, "math grade, n (%)" = 4))
```
```{r tm}
dat %>% 
    filter(mathgrade <= "(4) D or lower" &
               tvtime < "(9) Multiple response") %>% 
    group_by(tvtime, mathgrade) %>% 
    summarise(n = n()) %>% 
    mutate(pct = (n / sum(n) * 100) %>% round(1),
           `n (%)` = str_c(n, " (", pct, ")", sep = "")) %>% 
    select(mathgrade, tvtime, `n (%)`) %>% 
    spread(mathgrade, `n (%)`) %>% 
    kable(caption = "Tabulation of grades in math and TV time, percent total by TV time") %>% 
    kable_styling(full_width = FALSE,
                  position = "left") %>% 
    add_header_above(header = c(" " = 1, "math grade, n (%)" = 4))

```

<h2>4</h2>
__Create at least one graph from data in the data frame.__

Figure \@ref(fig:mathgrade) shows the count of subjects by math grades and frequency of having trouble getting homework done. For those who never had a problem getting homework done, grade A was most prevalent. But for all other groups, grade B was most prevalent. The most prevalent group had trouble completing their homework "just a few times a week."

```{r mathgrade, fig.cap="Math grades by frequency of trouble getting homework done"}
dsum <- dat %>% 
    filter(mathgrade <= "(4) D or lower" &
               homework < "(9) Multiple response") %>% 
    group_by(homework, mathgrade) %>% 
    summarise(n = n())

ggplot(data = dsum, mapping = aes(x = mathgrade, y = n)) +
    geom_bar(stat = "identity") +
    facet_grid(~ homework) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
    
```



<h2>Source code</h2>
[csde502_2021_assignment07_answers.Rmd](csde502_2021_assignment07_answers.Rmd)

```{r comment=''}
cat(readLines(con = "csde502_2021_assignment07_answers.Rmd"), sep = '\n')
```
